---
title: "Toy Example"
author: "Jianhui Gao"
date: "2026-03-31"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(ggplot2)
library(pROC)
library(patchwork)

script_dir <- normalizePath(getwd())
data_dir <- file.path(script_dir, "Data")
```

# Simulation Setup

We consider four data-generating mechanisms. In all cases, we generate
\[
A_i \sim \mathrm{Bernoulli}(0.5), \qquad 
\mathbf{X}_i \sim \mathcal{N}(\mathbf{0}, I_5), \qquad 
\mathbf{W}_i \sim \mathcal{N}(\mathbf{0}, I_5)
\]

Let $\boldsymbol{\beta}_X, \boldsymbol{\beta}_W \in \mathbb{R}^5$ be unit-norm vectors, and define
\[
\eta_i = \frac{\mathbf{X}_i^\top \boldsymbol{\beta}_X + \mathbf{W}_i^\top \boldsymbol{\beta}_W}{\sqrt{2}},
\quad
\alpha_0 = -0.5, \ \alpha_1 = 0.5.
\]

\paragraph{(i) Ideal case.}
We define
\[
S_i = \mathrm{expit}(\alpha_{A_i} + \eta_i), 
\qquad 
Y_i \mid S_i \sim \mathrm{Bernoulli}(S_i).
\]
In this case,
\[
S_i = \mathbb{E}[Y_i \mid \mathbf{X}_i, \mathbf{W}_i, A_i],
\quad 
\mathbb{E}[Y_i \mid S_i, A_i, \mathbf{W}_i] = S_i,
\]
so $S_i$ is a perfectly calibrated risk score.

\paragraph{(ii) S misspecified but E[Y|S,A] lies within the beta-calibration family and $\mathbb{E}[Y \mid S,A] = \mathbb{E}[Y \mid S,A,W]$.}
We define
\[
S_i = \eta_i, 
\qquad 
Y_i \mid S_i, A_i \sim \mathrm{Bernoulli}\big(\mathrm{expit}(\alpha_{A_i} + S_i)\big).
\]
Here,
\[
\mathbb{E}[Y_i \mid S_i, A_i, \mathbf{W}_i] = \mathbb{E}[Y_i \mid S_i, A_i],
\]
so $S_i$ is sufficient for predicting $Y_i$, but is not calibrated.

\paragraph{(iii) S misspecified and E[Y|S,A] doesnt lie within the beta-calibration family and $\mathbb{E}[Y \mid S,A] = \mathbb{E}[Y \mid S,A,W]$.}
We define
\[
S_i = \eta_i, 
\qquad
Y_i \mid S_i, A_i \sim \mathrm{Bernoulli}\big(\mathrm{expit}(\alpha_{A_i} + \gamma(S_i + 0.5 S_i^2))\big),
\]
with $\gamma=1$. In this case,
\[
\mathbb{E}[Y_i \mid S_i, A_i, \mathbf{W}_i] = \mathbb{E}[Y_i \mid S_i, A_i],
\]
but the regression function is nonlinear in $S_i$, so beta calibration are misspecified.

\paragraph{(iv) Additional information in $\mathbf{W}$.}
We define
\[
S_i = \mathbf{X}_i^\top \boldsymbol{\beta}_X, 
\qquad 
Y_i \mid S_i, \mathbf{W}_i, A_i \sim 
\mathrm{Bernoulli}\big(\mathrm{expit}(\alpha_{A_i} + S_i + \mathbf{W}_i^\top \boldsymbol{\beta}_W)\big).
\]
Here,
\[
\mathbb{E}[Y_i \mid S_i, A_i, \mathbf{W}_i] \neq \mathbb{E}[Y_i \mid S_i, A_i],
\]
so $\mathbf{W}_i$ provides additional predictive information beyond $S_i$.

# ROC Check
```{r}
draw_unit <- function(p) {
  z <- rnorm(p)
  z / sqrt(sum(z^2))
}

gen_data_ideal <- function(n = 2e4, p = 5) {
  beta_X <- draw_unit(p)
  beta_W <- draw_unit(p)
  
  A <- rbinom(n, 1, 0.5)
  X <- matrix(rnorm(n * p), n, p)
  colnames(X) <- paste0("X", 1:p)
  W <- matrix(rnorm(n * p), n, p)
  colnames(W) <- paste0("W", 1:p)
  
  eta <- (X %*% beta_X + W %*% beta_W) / sqrt(2)
  
  alpha <- c(-0.5, 0.5)
  gamma <- 1
  
  S <- plogis(alpha[A + 1] + gamma * eta)
  Y <- rbinom(n, 1, S)
  
  data.frame(Y, A, S, X, W)
}

gen_data_sufficient <- function(n = 2e4, p = 5) {
  beta_X <- draw_unit(p)
  beta_W <- draw_unit(p)
  
  A <- rbinom(n, 1, 0.5)
  X <- matrix(rnorm(n * p), n, p)
  colnames(X) <- paste0("X", 1:p)
  W <- matrix(rnorm(n * p), n, p)
  colnames(W) <- paste0("W", 1:p)
  
  S <- (X %*% beta_X + W %*% beta_W) / sqrt(2)
  
  alpha <- c(-0.5, 0.5)
  gamma <- 1
  
  prob <- plogis(alpha[A + 1] + gamma * S)
  Y <- rbinom(n, 1, prob)
  
  data.frame(Y, A, S, X, W)
}

gen_data_beta_fail <- function(n = 2e4, p = 5) {
  beta_X <- draw_unit(p)
  beta_W <- draw_unit(p)
  
  A <- rbinom(n, 1, 0.5)
  X <- matrix(rnorm(n * p), n, p)
  colnames(X) <- paste0("X", 1:p)
  W <- matrix(rnorm(n * p), n, p)
  colnames(W) <- paste0("W", 1:p)
  
  S <- (X %*% beta_X + W %*% beta_W) / sqrt(2)
  
  alpha <- c(-0.5, 0.5)
  gamma <- 1
  
  prob <- plogis(alpha[A + 1] + gamma * (S + 0.5*S^2))
  Y <- rbinom(n, 1, prob)
  
  data.frame(Y, A, S, X, W)
}

gen_data_w_extra <- function(n = 2e4, p = 5) {
  beta_X <- draw_unit(p)
  beta_W <- draw_unit(p)
  
  A <- rbinom(n, 1, 0.5)
  X <- matrix(rnorm(n * p), n, p)
  colnames(X) <- paste0("X", 1:p)
  W <- matrix(rnorm(n * p), n, p)
  colnames(W) <- paste0("W", 1:p)
  
  S <- X %*% beta_X
  
  alpha <- c(-0.5, 0.5)
  gamma <- 1
  delta <- 1
  
  prob <- plogis(alpha[A + 1] + gamma * S + delta * (W %*% beta_W))
  Y <- rbinom(n, 1, prob)
  
  data.frame(Y, A, S, X, W)
}
set.seed(1)

gen_list <- list(
  "Scenario 1" = gen_data_ideal,
  "Scenario 2" = gen_data_sufficient,
  "Scenario 3" = gen_data_beta_fail,
  "Scenario 4" = gen_data_w_extra
)

par(mfrow = c(2, 2))

for (nm in names(gen_list)) {
  dat <- gen_list[[nm]]()
  
  roc_obj <- roc(response = dat$Y, predictor = as.numeric(dat$S), quiet = TRUE)
  
  plot(
    roc_obj,
    main = paste0(nm, "\nAUC = ", round(auc(roc_obj), 3))
  )
}
```

# Calibration Check

```{r}
PlotCalib <- function(y, p, A, main = "") {
  
  plot(NULL,
       xlim = c(0,1), ylim = c(0,1),
       xlab = "Predicted probability",
       ylab = "Observed probability",
       main = main)
  
  abline(0,1, lty = 2, col = "grey70")
  
  # smooth calibration
  fit0 <- loess(y[A==0] ~ p[A==0], span = 0.75)
  fit1 <- loess(y[A==1] ~ p[A==1], span = 0.75)
  
  grid <- seq(0,1,length.out=200)
  
  lines(grid, predict(fit0, newdata = grid), lwd = 2)
  lines(grid, predict(fit1, newdata = grid), lwd = 2, lty = 2)
  
  legend("topleft",
         legend = c("A=0","A=1"),
         lwd = 2, lty = c(1,2),
         bty = "n")
}

gen_list <- list(
  "Scenario 1" = gen_data_ideal,
  "Scenario 2" = gen_data_sufficient,
  "Scenario 3" = gen_data_beta_fail,
  "Scenario 4" = gen_data_w_extra
)

par(mfrow = c(2,2), mar = c(4,4,2,1))

for (nm in names(gen_list)) {
  
  dat <- gen_list[[nm]]()
  
  # choose scale
  p_use <- if (nm == "Scenario 1") dat$S else plogis(dat$S)
  
  PlotCalib(dat$Y, p_use, dat$A, main = nm)
}

```


# Point Estimate

```{r, warning=FALSE, message=FALSE}
plist <- list()
relist <- list()

for(i in 1:4){
  results <- readRDS(file.path(data_dir, paste0("toy", i, ".rds")))
  results_df <- do.call(rbind, results)

# Oracle truth
oracle_mean <- results_df %>%
  filter(method == "oracle") %>%
  summarise(val = mean(estimate)) %>%
  pull(val)

left_df <- results_df %>%
  filter(method != "oracle") %>%
  mutate(
    method = factor(
      method,
      levels = c("supervised", "naive", "beta calibration",
                 "kernel smoothing", "poly(S)", "poly(S+W)",
                 "spline(S)", "spline(S+W)")
    )
  )

p1_boxplot <- left_df %>% 
  filter(!is.na(estimate)) %>% 
  ggplot(aes(x = method, y = estimate)) +
  geom_boxplot(width = 0.5, outlier.size = 1.5) + 
  geom_hline(
    yintercept = oracle_mean,
    linetype = "dashed",
    linewidth = 0.8,
    color = "red"
  ) +
  labs(
    title = paste0("Scenario ", i),
    x = NULL,
    y = expression(E[Y~"|"~A==1] - E[Y~"|"~A==0])
  ) +
  theme_bw(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 20, hjust = 1)
  )

right_df <- results_df %>%
  filter(!method %in% c("oracle")) %>%
  na.omit() %>%
  mutate(
    method = factor(
      method,
      levels = c("supervised", "naive", "beta calibration",
                 "kernel smoothing", "poly(S)", "poly(S+W)",
                 "spline(S)", "spline(S+W)")
    )
  )

re_df <- right_df %>%
  filter(estimate < Inf & estimate > -Inf) %>%
  mutate(sq_error = (estimate - oracle_mean)^2) %>%
  group_by(method) %>%
  summarise(mse = mean(sq_error), .groups = "drop")

mse_supervised <- re_df %>%
  filter(method == "supervised") %>%
  pull(mse)

re_df <- re_df %>%
  mutate(RE = mse_supervised / mse)

p2 <- ggplot(re_df %>%
               filter(!method %in% c("naive", "supervised", "beta calibration"))
               , aes(x = method, y = RE)) +
  geom_col(width = 0.65, fill = "grey80", color = "black") +
  geom_hline(
    yintercept = 1,
    linetype = "dashed",
    linewidth = 0.8,
    color = "red"
  ) +
  labs(
    title = paste0("Scenario ", i),
    x = NULL,
    y = "RE (MSE)"
  ) +
  theme_bw(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 20, hjust = 1)
  )

plist[[i]] <- p1_boxplot
relist[[i]] <- p2


}
library(patchwork)
(plist[[1]] + plist[[2]] + plist[[3]] + plist[[4]]) +
  plot_annotation(
    theme = theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5))
  )
```

# Relative Efficiency

```{r}
(relist[[1]] + relist[[2]] + relist[[3]] + relist[[4]]) +
  plot_annotation(
    theme = theme(plot.title = element_text(face = "bold", size = 16, hjust = 0.5))
  )
```
